---
title: "Dataframes - In Class lab"
author: "Tyler Guerrero"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Working with loaded in csv files


Use the `read.csv()` function to read the data from the `gates_donations.csv` file (available on Canvas, make sure you save it in the project repository folder) into a variable called `grants` 

```{r}
## if you save it in the same folder as this file this should work. 
## But you may need edit the string to reflect the actual file name: 

grants <- read_csv("gates_donations.csv")


head(grants)
```

Because you're in an RProject, you don't need to set your working directory in RStudio! The working directory is automatically set to the repository location. (You can use `getwd()` to confirm this)

- Use the glimpse function to look at the loaded data

```{r}
glimpse(grants)

```


- Create a variable `org` that contains the `organization` column of the dataset

```{r}
org <- grants$organization
org
```


- Confirm that the "organization" column is a vector using the `is.vector()` function. (This is a useful debugging tip if you hit errors later!)

```{r}
is.vector(org)

```


Now you can ask some interesting questions about the dataset:

- What was the mean grant value?

```{r}
total <- grants$total_amount
mean(total)
```


- What was the dollar amount of the largest grant?

```{r}
max(total)
```


- What was the dollar amount of the smallest grant?

```{r}
min(total)
```


- Which organization received the largest grant?

```{r}
largest_grant <- which.max(total)
org[largest_grant]

```


- Which organization received the smallest grant?

```{r}
smallest_grant <- which.min(total)
org[smallest_grant]
```

- How many grants were awarded in 2010?

```{r}
nrow(grants[grants$start_year == 2010, ])
```

- How much was awarded in 2010? 

```{r}
sum(grants[grants$start_year == 2010, "total_amount"])

```


# Working with built-in data sets

Load R's "USPersonalExpenditure" dataset using the `data()` function

This will produce a data frame called `USPersonalExpenditure`
```{r}
data("USPersonalExpenditure")
```

The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using) 

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
USPE <- data.frame(USPersonalExpenditure)
USPE
```


What are the current column names of your dataframe?
```{r}
colnames(USPE)
```


Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?
```{r}
rownames(USPE)
```


Add a column "Category" to your data frame that contains the rownames

```{r}
USPE <- rownames_to_column(USPE, var = "Category")
```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy. 

```{r}
USPE <- pivot_longer(USPE,
                     cols = c("X1940", "X1945", "X1950", "X1955", "X1960"),
                     names_to = "Year",
                     values_to = "Amount_$")
USPE
```


How much money was spent on personal care in 1940?
```{r}
sum(USPE$'Amount_$'[USPE$Category == "Personal Care" & USPE$Year == "X1940"])
```


How much money was spent on Food and Tobacco in 1960?
```{r}
sum(USPE$'Amount_$'[USPE$Category == "Food and Tobacco" & USPE$Year == "X1960"])

```


What was the highest expenditure category in 1960?
```{r}
subset_1960 <- USPE[USPE$Year == "X1960", ]
subset_1960$Category[which.max(subset_1960$`Amount_$`)]

```


Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year
```{r}
lowest_category <- function(year){
  year_value <- paste0("X",year)
  
  that_year_data <- USPE[USPE$Year == year_value, ]
  that_year_data$Category[which.max(that_year_data$'Amount_$')]
}

lowest_category(1950)
```

```{r}
year_vec <- seq(1940, 1960, by = 5)
sapply(year_vec, lowest_category)
```


Using your function, determine the lowest spending category of each year
Hint: use the `sapply()` function to apply your function to a vector of years

```{r}
USPE$Year_num <- as.numeric(str_sub(USPE$Year, start = 2, end = 5))

sapply(USPE$Year_num, lowest_category)
```
